Search for Spatially Extended \fermi-LAT Sources Using Two Years of Flight Data

### 2011-08-26 extended_catalog/paper/extended_catalog.tex
updated test(?)
alexs introduction changes

\newif\ifcolorfigure
%\colorfiguretrue
\colorfigurefalse

@@

\begin{abstract}
We present a new method for quantifying the spatial extension of sources
with the Large Area Telescope (LAT), the primary science instrument on
the {\em \fermi Gamma-ray Space Telescope} (\fermi).  Spatial extension
is an important characteristic for correctly associating LAT sources with
their counterparts at other wavelengths. We perform a series of Monte
Carlo simulations to validate this tool and calculate the LAT's threshold
for detecting the spatial extension of sources.  We then 
test all sources in the two-year source catalog for extension. We
report the detection of nine spatially extended sources in addition to
the twelve spatially extended sources reported in the second \fermi-LAT
Catalog (2FGL).
\end{abstract}

@@

A number of astrophysical sources classes including supernova remnant
(SNRs), pulsar wind nebulae (PWNe), molecular clouds, normal galaxies,
galaxy clusters could be expected to spatially resolvable at \gev
energies.  Dark matter satellites are also hypothesized to be seen
as spatially extended at \gev energies.  The LAT Collaboration has
previously reported on five SNRs which are spatially extended beyond
the LAT PSF at \gev energies (IC443, W28, W44, W51C, and RX\,J1713.7$-$3946
\citep{ic443,w28,w44,w51c,rx_j1713_lat}). In addition, three extended PWNe
were detected as being extended: MSH\,15$-$52, Vela~X, and HESS\,J1825$-$137
\citep{msh1552,velax,fermi_hess_j1825}. Two close-by galaxies the Large
and Small Magalenic Clouds and one radio galaxy Centarus A were spatially
resolved at \gev energies \citep{lmc,smc,cen_a_lat}.  Furthermore,
a number of other sources previously detected at \gev energies are
positionally coincident with sources that exhibit extension at other
wavelengths and could be expected to be spatially resolvable at \gev
energies.

At \tev energies many sources have been recently detected as
begin spatially extended using air Cherenkov detectors.  Most
prominent was a survey of the Galactic plane using the High Energy
Stereoscopic System (H.E.S.S) which reported 14 spatially extended
sources with extensions varying from $\sim0\fdg1$ to $\sim0\fdg25$
\citep{hess_plane_survey}.  In fact, within our Galaxy only very few
sources detected at \tev energies (most notably the $\gamma$-ray binaries
LS\,5039 \citep{HESSLS5039}, LS I+61$-$303 \citep{MAGICLSI, VERITASLSI}
HESS\,J0632+057 \citep{HESS0632}, and the Crab nebula \citep{crab_weekes})
have no significant extension detected.  High energy $\gamma$-rays from
these sources are produced by the decay of $\pi^0$s produced by hadronic
interactions with interstellar matter and by relativistic electrons
due to Inverse Compton (IC) scattering and Bremsstrahlung radiation
\citep{blandford_and_eichler_1987}.  It is likely that the \gev and
\tev emission from these sources originates from the same population of
high-energy particles and so at least some of these sources should be
detected as extended sources at \gev energies.  Studying 
these \tev sources at \gev energies would help to
determine the emission mechanisms producing these high energy photons.

The Large Area Telescope (LAT) is a pair conversion telescope on the
{\em \fermi Gamma-ray Space Telescope} (\fermi) that been surveying the
$\gamma$-ray sky since 2008 June.  The LAT has broad energy coverage
(20 \mev to $>300$ \gev), wide field of view ($\sim 2.4 \sr$), and large
effective area ($\sim 8000\ \cm^2$ at $>1 \gev$).  
Using one year of all-sky surveying data, the LAT Collaboration published
a catalog of 1451 sources significantly detected at \gev energies called 1FGL \citep{first_cat}.
Using two years of data, a second catalog called 2FGL reported on 1873
sources \citep{second_cat}.
The counterparts of many of these sources can be spatially resolved
when observed at other frequencies but detecting the spatial extension
of these sources at \gev energies is difficult because the size of the
point-spread function (PSF) of the LAT is comparable to the typical size
of many of these sources.

The capability to spatially resolve \gev $\gamma$-ray
sources is important for several reasons.  First, multiple potential
counterparts detected at other frequencies exist within the localization
confidence region of a typical LAT source along the Galactic plane.
Finding a coherent source extension across different energy bands can
help to associate a LAT source to an otherwise confused counterpart.
Furthermore, dark matter satellite in our galaxy have been predicted to be
detected and spatially extended at \gev energies and could be discovered
outside the galactic plane \citep{pre_luanch_dark_matter_fermi}.  Also,
due to the strong energy dependence of the LAT PSF, the spatial and
spectral information about a source do not simply decouple. An inaccurate
spatial model will bias the spectral model of the source. Specifically,
modeling a spatially extended source as point-like will systematically
shift a spectral analysis to softer indices. Furthermore, correctly
modeling a source's extension is important for improving the model of
the sky and removing excess residuals, for example in the region around
surrounding the Large Magellanic Clouds \citep{first_cat}.  Such excess
residuals potentially bias the significance and measured spectrum of
neighboring sources in the densely populated Galactic plane.

Each of the previously detected extended LAT sources was spatially
resolved by a dedicated analysis of the particular source so
we expect that a systematic scan of all LAT detected sources
could uncover additional spatially extended sources.  For these
reasons, in Section~\ref{analysis_methods_section} we present
a new method for analyzing spatially extended LAT sources. In
Section~\ref{monte_carlo_validation} we demonstrate that this tool
can be used for an unbiased statistical test of the extension of
LAT sources and in Section~\ref{extension_sensitivity} we calculate
the LAT's detection threshold to resolve the extension of a source.
In Section~\ref{extended_source_search_method} we describe a search for
new spatially extended LAT sources and in Section~\ref{validate_known}
we reanalyzing the twelve previously detected extended sources
included in 2FGL.  In Section~\ref{test_2lac_sources} we further
demonstrate that our detection method does not misidentify
point sources as being extended 
by testing the extension of Active
galactic nucleus (AGN) believed to be unresolvable. Finally, in
Section~\ref{new_ext_srcs_section} we present the detection of the
extension of nine spatially extended sources that were reported in 2FGL
but not previously spatially resolved.

>>>>

\newif\ifcolorfigure
\colorfiguretrue
%\colorfigurefalse

@@

\begin{abstract}
Spatial extension is an important characteristic for correctly
associating LAT sources with their counterparts at other wavelengths
and for obtaining an unbiased model of their spectra.  We present a new
method for quantifying the spatial extension of sources with the Large
Area Telescope (LAT), the primary science instrument on the {\em \fermi
Gamma-ray Space Telescope} (\fermi).  We perform a series of Monte Carlo
simulations to validate this tool and calculate the LAT's threshold for
detecting the spatial extension of sources.  We then test all sources
in the second \fermi-LAT catalog (2FGL) for extension. We report the
detection of nine spatially extended sources in addition to the twelve
spatially extended sources reported in 2FGL.
\end{abstract}

@@

A number of astrophysical source classes including supernova remnants
(SNRs), pulsar wind nebulae (PWNe), molecular clouds, normal galaxies,
and galaxy clusters are expected to be spatially resolvable at
\gev energies.  Dark matter satellites are also hypothesized to
be spatially extended.  The Large Area Telescope
(LAT) on the {\em \fermi Gamma-ray Space Telescope} (\fermi) has
detected five SNRs which are significantly extended
at \gev energies: IC443, W28, W44, W51C, and RX\,J1713.7$-$3946
\citep{ic443,w28,w44,w51c,rx_j1713_lat}. In addition, three extended
PWNe were detected as being extended: MSH\,15$-$52, Vela~X, and
HESS\,J1825$-$137 \citep{msh1552,velax,fermi_hess_j1825}. Two
close-by galaxies, the Large and Small Magalenic Clouds, and one
radio galaxy, Centarus A, were spatially resolved at \gev energies
\citep{lmc,smc,cen_a_lat}.  A number of additional sources detected
at \gev energies are positionally coincident with sources that exhibit
extension at other wavelengths. These sources could be expected to be spatially
resolvable by the LAT at \gev energies.

The current generation of air Cherenkov detectors have
made it apparent that many sources can be spatially resolved
at even higher energies.
Most
prominent was a survey of the Galactic plane using the High Energy
Stereoscopic System (H.E.S.S) which reported 14 spatially extended
sources with extensions varying from $\sim0\fdg1$ to $\sim0\fdg25$
\citep{hess_plane_survey}.  In fact, within our Galaxy only very few
sources detected at \tev energies (most notably the $\gamma$-ray binaries
LS\,5039 \citep{HESSLS5039}, LS I+61$-$303 \citep{MAGICLSI, VERITASLSI}
HESS\,J0632+057 \citep{HESS0632}, and the Crab nebula \citep{crab_weekes})
have no detectable extension.  High-energy $\gamma$-rays from
these sources are produced by the decay of $\pi^0$s produced by hadronic
interactions with interstellar matter and by relativistic electrons
due to Inverse Compton (IC) scattering and Bremsstrahlung radiation
\citep{blandford_and_eichler_1987}.  It is likely that the \gev and
\tev emission from these sources originates from the same population of
high-energy particles and so at least some of these sources should be
detected at \gev energies.  Studying 
these \tev sources at \gev energies would help to
determine the emission mechanisms producing these high energy photons.

The LAT is a pair conversion telescope that been surveying the
$\gamma$-ray sky since 2008 June.  The LAT has broad energy coverage
(20 \mev to $>300$ \gev), wide field of view ($\sim 2.4 \sr$), and large
effective area ($\sim 8000\ \cm^2$ at $>1 \gev$).  
Using one year of all-sky surveying data, the LAT Collaboration published
a catalog of 1451 sources significantly detected at \gev energies called 1FGL \citep{first_cat}.
Using two years of data, a second catalog called 2FGL reported on 1873
sources \citep{second_cat}.
The counterparts of many of these sources can be spatially resolved
when observed at other frequencies but detecting the spatial extension
of these sources at \gev energies is difficult because the size of the
point-spread function (PSF) of the LAT is comparable to the typical size
of many of these sources.

The capability to spatially resolve \gev $\gamma$-ray
sources is important for several reasons.  
Finding a coherent source extension across different energy bands can
help to associate a LAT source to an otherwise confused counterpart.
Furthermore, some of the dark matter substructure in our Galaxy 
is predicted to be spatially extended at \gev energies \citep{pre_luanch_dark_matter_fermi}.  
Characterization of spatial extension could help to identify this substructure.
Also,
due to the strong energy dependence of the LAT PSF, the spatial and
spectral character of a source do not decouple. An inaccurate
spatial model will bias the spectral model of the source and vice versa. Specifically,
modeling a spatially extended source as point-like will systematically
shift a spectral analysis to softer indices. Furthermore, correctly
modeling a source's extension is important for 
understanding an entire region of the sky. For example,
an incorrect model of the spatially extended LMC introduced
significant residuals in the surrounding region \citep{first_cat,second_cat}.
Such residuals can bias the significance and measured spectrum of
neighboring sources in the densely populated Galactic plane.

Previous analysis of
extended LAT sources were performed as dedicated studies
of individual sources, so we expect
that a systematic scan of all LAT detected sources
could uncover additional spatially extended sources.  For these
reasons, in Section~\ref{analysis_methods_section} we present
a new systematic method for analyzing spatially extended LAT sources. In
Section~\ref{monte_carlo_validation} we demonstrate that this method
can be used to test the statistical significance of
the extension of
a LAT sources and in Section~\ref{extension_sensitivity} we calculate
the LAT's detection threshold to resolve the extension of a source.
In Section~\ref{extended_source_search_method} we describe a search for
new spatially extended LAT sources and in Section~\ref{validate_known}
we reanalyze the twelve previously detected extended sources
included in 2FGL.  In Section~\ref{test_2lac_sources} we further
demonstrate that our detection method does not misidentify
point sources as being extended 
by testing the extension of active
galactic nuclei (AGN) believed to be unresolvable. Finally, in
Section~\ref{new_ext_srcs_section} we present the detection of the
extension of nine spatially extended sources that were reported in 2FGL
but not previously spatially resolved.
